---
title: Build a letter recognition classifier
author: ''
date: '2017-12-28'
description: "Develop a model that will predict the hand-written capital letters (English) displayed in a picture"
categories:
  - R
tags:
  - classification
  - clustering
  - k-NN
  - SVD
---

In this analysis i’ll develop a model that will predict the hand-written capital letters (English) displayed
in a picture, using various of it's attributes.


# INTRODUCTION

**Dataset Information**

This data set was originally used in the paper of **P. W. Frey and D. J. Slate."Letter Recognition Using Holland-style Adaptive Classifiers"**
(Machine Learning Vol 6 #2 March 91). It investigated the ability of several
variations of Holland-style adaptive classifier systems to learn to
correctly guess the letter categories associated with vectors of 16
integer attributes extracted from raster scan images of the letters. The actual
data can be found at Letter Recognition Data Set Machine Learning Repository:
https://archive.ics.uci.edu/ml/datasets/Letter+Recognition


The objective is to identify each of a large number of 
black-and-white rectangular pixel displays as one of the 26 capital letters in 
the English alphabet. The character images were based on 20 different fonts and 
each letter within these 20 fonts was randomly distorted to produce a file of 
unique stimuli. Each stimulus was converted into primitive numerical 
attributes (statistical moments and edge counts) which were then scaled to fit 
into a range of integer values from 0 through 15.
So in total there are **20000 observations** of **17 variables**. There are no 
missing values in the data set.

**The variables**

1.	lettr	capital letter	(26 values from A to Z) 
2.	x-box	horizontal position of box	(integer) 
3.	y-box	vertical position of box	(integer) 
4.	width	width of box	(integer) 
5.	high height of box	(integer) 
6.	onpix	total # on pixels	(integer) 
7.	x-bar	mean x of on pixels in box	(integer) 
8.	y-bar	mean y of on pixels in box	(integer) 
9.	x2bar	mean x variance	(integer) 
10.	y2bar	mean y variance	(integer) 
11.	xybar	mean x y correlation	(integer) 
12.	x2ybr	mean of x * x * y	(integer) 
13.	xy2br	mean of x * y * y	(integer) 
14.	x-ege	mean edge count left to right	(integer) 
15.	xegvy	correlation of x-ege with y	(integer) 
16.	y-ege	mean edge count bottom to top	(integer) 
17.	yegvx	correlation of y-ege with x	(integer)


# PRE-PROCCESSING 

Since the original data set didn't contain any names on the variables, after i imported
the data in R environment, i assigned the appropriate variable names. That was completed
with the help of **letter-recognition.names** text file which accompanied the 
original data.

During the development of the prediction model, in order to predict the letter on 
new observations, at first we've used all variables. Then we excluded some of the 4 
correlated variables and finally we've used the variables created from the SVD 
algorithm instead of the original.

Before we apply the hierarchical clustering, we created a data set with each
letter as one observation and the average of each variable, as a distinct 
variable **(named letter.cluster)**.

```{r PRE-PROCCESSING, echo=TRUE, message=FALSE, warning=FALSE}
# Load Libraries
library(tidyverse)
library(stringr)
library(scales)
library(ggthemes)
library(corrplot)
library(caret)
library(gmodels)
library(class)
library(ggdendro)


# DATA WRANGLING ###############################################################
# Insert dataset
letter <- read.table("/Users/manos/OneDrive/Projects/R/Data/letter-recognition.data",
                  sep = ",")

# Create vector with new variable names
name <- c("lettr", "x-box", "y-box", "width", "high", "onpix", "x-bar", "y-bar", 
          "x2bar", "y2bar", "xybar", "x2ybr", "xy2br", "x-ege", "xegvy", "y-ege",
          "yegvx")

# Change the variable names of the dataset
names(letter) <- name

# Make it a tibble
letter <- as_data_frame(letter)

# Check for missing values
sum(is.na(letter))

# Create the summary dataset
letter.cluster <- letter %>% 
  group_by(lettr) %>% 
  summarise_each(funs(mean))

```


# EXPLORATORY ANALYSIS

In order to have an overview of the data-set's variability, i produced a box-plot
matrix of all variables and each letter on the x-axis

```{r exploratory box-plot, echo=TRUE, fig.height=10, message=FALSE, warning=FALSE}
# Plot a box-plot for each letter and variable
letter %>% 
  gather("type", "n", 2:17) %>% 
  ggplot()+
  geom_boxplot(aes(x = lettr, y = n), outlier.size = .5, fill = "steelblue2", alpha = .7)+
  facet_grid(type ~.)+
  labs(y = "", x = "", 
       title = "Box-Plots of all variables for each letter", 
       subtitle = "")+
 theme_fivethirtyeight()

```

It looks that in some variables, the letters levels have significant different 
variance, than others. In particular, in **y-box** and **high** variables the different 
letters don't seem to have any significant differences. On the other hand in **x-bar**,
 **x2ybr**, **y-bar**, **y-ege** variables the different letters seem to have large
 differences.


We should check for correlated variables in the data set. Below you can see a
correlogram of the data set variables.

```{r exploratory correlogram, echo=TRUE, message=FALSE, warning=FALSE}
# Plot a correlogram to check for correlations
corrplot(cor(letter[,2:17]), method="number", type = "lower", number.cex = .6)

```

It seems that the first 4 variables **(x-box, y-box, width, high)** are quite 
correlated. Especially x-box & width (0.85) and y-box & high (0.82) are highly 
correlated.

# MODELLING

An interesting insight that could be obtained from the data set is to try to 
predict the letter based on the variables produced by the digitized image. 
Classification is the problem of identifying to 
which of a set of categories (sub-populations) a new observation belongs, on the 
basis of a training set of data containing observations (or instances) whose 
category membership is known. So we must develop a model that classifies 
(categorize) every observation (case) to one of the 26 letters of the alphabet.

## **K-Nearest Neighbors (k-NN)**

The K-nearest neighbors (kNN) algorithm used, in order to classify observations
in a certain category (letter) by using the rest of the variables in the data set.
It begins with a training data set made up of examples that are
classified into several categories, as labeled by a nominal variable. Assume that we
have a test data set containing unlabeled examples that otherwise have the same
features as the training data. For each record in the test data set, kNN identifies k
records in the training data that are the "nearest" in similarity, where k is an integer
specified in advance. The unlabeled test instance is assigned the class of the majority
of the k-nearest neighbors. We splitted the data set into a training data set
(containing 70 % of the original data set observations) in which we build the 
classification model and a testing data set (containing the rest of the original
data set observations - 30%) in which we tested the model we build before.
The results are presented in the "results" section. 

```{r KNN, echo=TRUE, message=FALSE, warning=FALSE}

# Create a vector with the 70% of the dataset with respect to letter
set.seed(10)
inTrain = createDataPartition(letter$lettr, p = .7)[[1]]

# Assign the 70% of observations to training data
training <- letter[inTrain, -1]
training.lettr <- c(t(letter[inTrain, 1]))

# Assign the remaining 30 % of observations to testing data
testing <- letter[-inTrain, -1]
testing.lettr <- c(t(letter[-inTrain, 1]))


# Run knn algorithm on training dataset
# Create the knn model
knn_model <- knn(train = training, test = testing, cl = training.lettr, k = 3)

# Create a table in order to check the performance of the classification model
t <- CrossTable(x = testing.lettr,y = knn_model,
           prop.chisq=FALSE)

# Calculate prediction success rate for all letters
mean(diag(t$prop.col))

# Check prediction success rate for each letter
diag(t$prop.col)

```

The accuracy obtained from the nearest neighbors classification algorithm is around 
**94,8%**.  I've used 3 nearest neighbors (k=3) for building the final model. Alternative
number of neighbors (4,5,6,7 etc) produced slightly lower accuracy rate (94.3% - 94,7%).
The model is better at predicting some letters, such as L, M, U, Z with more than 
98% accuracy rate. On the other hand, other letters such as H, R, B are more difficult
to predict (<90% accuracy rate). In the plot below you can see each letter accuracy
rate of the final model.

```{r echo=TRUE, message=FALSE, warning=FALSE}
# Create a data set with the accuracy rate of each letter
letters <- diag(t$prop.col) %>% 
  as.data.frame()

# Make the appropriate transformations in order to plot
letters$letter <- rownames(letters)

# Plot the accuracy rate per letter
ggplot(letters)+
  geom_bar(aes(x = letter, y = .), stat="identity", fill = "steelblue2", alpha = .7)+
    scale_y_continuous(labels=percent) +
  labs(title = "Accuracy rate % per letter (K-NN algorithm)", 
       x = "", y = "Accuracy rate %", subtittle = "")+
 theme_fivethirtyeight()
  

```

## **Singular Value Decomposition (SVD)**

Since i discovered that there are some correlated variables in the data set, 
i applied the SVD algorithm in order to create a data set with uncorrelated 
variables. SVD is an algebraic tool that has many potential uses such as:  

* Dimensionality reduction
* Indexing (LSI)
* Visualization/clustering of high-dimensional objects
* Similarity computations/outlier detection
* Rule mining, treatment of missing/wrong values

Below i created a new dataset with the new SVD variables and build a new model
using k-NN algorithm.

```{r SVD, echo=TRUE, message=FALSE, warning=FALSE}
# Create an in Dataset 
letter.svd <- svd(letter[,2:17])

# Create a dataset with the SVD values
letter2 <- as_data_frame(cbind(letter[,1],letter.svd$u))

# Keep specific SVD variables
letter2 <- letter2[,1:8]

# Create a vector with the 70% of the dataset with respect to letter
set.seed(10)
inTrain = createDataPartition(letter2$lettr, p = .7)[[1]]

# Assign the 70% of observations to training data
training <- letter2[inTrain, -1]
training.lettr <- c(t(letter2[inTrain, 1]))

# Assign the remaining 30 % of observations to testing data
testing <- letter2[-inTrain, -1]
testing.lettr <- c(t(letter2[-inTrain, 1]))


# Run knn algorithm on training dataset
# Create the knn model
knn_model.svd <- knn(train = training, test = testing, cl = training.lettr, k = 3)

# Create a table in order to check the performance of the classification model
s <- CrossTable(x = testing.lettr,y = knn_model.svd,
           prop.chisq=FALSE)

# Calculate prediction success rate for all letters
mean(diag(s$prop.col))

# Check prediction success rate for each letter
diag(s$prop.col)

```

Even when all SVD variables (16) were used, the prediction rate is 
around 93.4% (lower than the original of 94.8%). When we use less (e.g. 8), the 
prediction rate is getting worst (83.4%). It seems that SVD does not help us improve 
the prediction model. Almost all original variables have variability (add information)
so they are useful for predicting the letter


## **Hierarchical clustering**
In this part of the analysis the objective is to assign the cases into 
clusters. Cluster analysis or clustering is the task of grouping a set of objects
in such a way that objects in the same group (called a cluster) are more similar 
(in some sense or another) to each other than to those in other groups (clusters).
Finally i build a hierarchical clustering model in order to understand which 
letters are "closer" to others and maybe misclassified. We applied it on the 
pre-processed data set (letter.cluster). 

```{r Clustering, echo=TRUE, message=FALSE, warning=FALSE}

# Tranform the summary dataset to a dataframe
letter.cluster <- as.data.frame(letter.cluster)

# Assign letters variable as rownames
rownames(letter.cluster) <- letter.cluster$lettr

# Run the hierarchical clustering 
clusters <- hclust(dist(letter.cluster))

```

At the plot below you can
see the hierarchical clustering output. When we compare these results with 
the misclassification rates of the original model, it seems that for some letters
like "B", "D", "F" it looks relevant. For other letters such "A", "C" don't match.


```{r echo=TRUE, message=FALSE, warning=FALSE}
# Plot the dendogram
ggdendrogram(clusters, rotate = TRUE, size = 2)+
  labs(title = "Dendogram of hierarchical clustering model")

```



# RESULTS

Finally, after testing various models using k-NN algorithm, the best model used 
the original predictors & 3 nearest neighbors (k=3). The accuracy obtained from 
from the final model is **94,8%**.
I also tried the **SVD technique** in order to check if the new uncorrelated variables
would help us improve the prediction accuracy rate of the letter in a new 
observation. But the model wasn't improved.


