---
title: Forecasting GDP in the eurozone
author: ''
date: '2018-08-26'
description: "Forecasting eurozone quarterly GDP (Gross domestic product) using Eurostat data & ARIMA modelling"
slug: forecasting-arima-modelling-gdp-eurostat
categories:
  - R
tags:
  - Forecasting
  - Arima
  - Eurostat
---


This analysis uses public Eurostat datasets, to forecast future total quarterly 
GDP of all eurozone countries. Eurostat is the statistical
office of the European Union situated in Luxembourg. Its mission is to provide 
high quality statistics for Europe. Eurostat offers a whole range of important 
and interesting data that governments, businesses, the education sector, 
journalists and the public can use for their work and daily life.  
In particular, the eurozone (EU 19) quarterly GDP (Gross domestic product) 
dataset is used. The eurozone consists of 19 countries: Austria, Belgium, Cyprus, 
Estonia, Finland, France, Germany, Greece, Ireland, Italy, Latvia, Lithuania, 
Luxembourg, Malta, the Netherlands, Portugal, Slovakia, Slovenia, and Spain. 

Gross domestic (GDP) is a monetary measure of the market value of all the final
goods and services produced in a period (quarterly or yearly) of time. It is 
commonly used to determine the economic performance of a country or regions.

The [Eurostat package](https://cran.r-project.org/web/packages/eurostat/index.html) 
used to obtain the dataset and [Forecast package](https://cran.r-project.org/web/packages/forecast/index.html) 
for the ARIMA modelling. 

More details about the ETL steps can be found, in the actual code, at the link 
at the end of the article.

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}

# Libraries
library(tidyverse)
library(rvest)
library(eurostat)
library(ggthemes)


```



```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}

# Download the dataset & label the data
dat <- get_eurostat("namq_10_gdp", time_format = "num")
dat <- label_eurostat(dat)

# Save the data
saveRDS(dat, file = "./data/dat.RDS")
```


```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}

# Load the dataset
dat <- readRDS("/Users/manos/OneDrive/Projects/R/All_Projects/2018_07_TimeSeriesForecast/data/dat.RDS")

library(forecast)

selected_geo <- "Euro area (19 countries)"

# Filter the quarterly EU19 GDP dataset
dat <- 
dat %>% 
  mutate(geo = as.character(geo)) %>% 
  filter(unit == "Current prices, million euro",
         s_adj == "Unadjusted data (i.e. neither seasonally adjusted nor calendar adjusted data)",
         na_item == "Gross domestic product at market prices",
         geo %in% selected_geo) %>% 
  arrange(time)


# Create the time series object
gdp_ts <- ts(dat[, 6], start = c(1995, 1), frequency = 4)

```


# Exploratory Analysis

During exploratory analysis, we try to discover patterns in the time series such as:  
- **Trend** A pattern involving long-term increase or decrease in the time series  
- **Seasonality**  A period pattern exists due to the calendar (e.g. quarter,
month, weekday)  
- **Cyclicity** A pattern exists where the data exhibits rise and fall that 
are not of a fixed period (duration usually of at least two years)  

Below there is a time series plot of the Eurozone countries quarterly GDP since 1995.

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

# Simple time series plot
autoplot(gdp_ts, facets = TRUE)+
  labs(title = "Quarterly GDP for Eurozone countries",
       subtitle = "In million euro (€)",
       y = "Quarterly GDP (€)",
       x = "year") +
  geom_smooth(method = "loess") +
  scale_x_continuous(breaks = seq(1995, 2018, 1)) +
  theme_fivethirtyeight() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

There are a few outputs from the time series plot above:   
- We can say that there is an overall positive trend  
- There is no noticeable increased/decreased variability in the trend  
- It looks that there is some seasonality, but needs further investigation  
- There is no cyclicity  
- There is a significant disruption in GDP growth around years 2008-9


A seasonal plot is used below to investigate seasonality. A seasonal plot is 
similar to a time plot except that the data are plotted against the individual 
“seasons” in which the data were observed.

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
## Seasonal plot
ggseasonplot(gdp_ts) +
  labs(title = "Seasonal GDP plot for Eurozone countries",
        subtitle = "In million euro (€)",
       y = "GDP (€)")+
  theme_fivethirtyeight()
```

This strengthens our confidence for seasonality in the time series. It seems that the
4th quarter is always the higher, while the 1st is the lowest in each year.
The 2nd & 3rd are about the same. 


A lag plot will help us understand if there is autocorrelation in the time series.
Another way to look at time series data is to plot each observation against another
observation that occurred some time previously. For example, you could plot yt 
against yt−1. This is called a lag plot because you are plotting the time series 
against lags of itself. 
Basically it is a scatterplot between the time series and the lagged values of the time series. 



```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
## Lag plot of time series
gglagplot(gdp_ts)+
    labs(title = "Lag plot of quarterly GDP for Eurozone countries",
        subtitle = "In million euro (€)",
       y = "GDP (€)")+
  theme_fivethirtyeight()
```

There is a strong seasonality at lag 4 (1 year), as all quarters line plots follow
an almost identical path.

Below there is an autocorrelation function plot. The correlations associated with
the lag plots form what is called the autocorrelation function (ACF). 
Spices that exceeds the confidence intervals (blue lines) indicates that autocorrelation 
with specific lag is statistically significant (different than zero)


```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Autocorellation plot
ggAcf(gdp_ts)+
  ggtitle("Autocorellation function plot") +
  theme_fivethirtyeight()
```

It looks that there are significant autocorrelations at all lags, which indicates a 
trend and/or seasonality in the time series.


We can also use the Ljung-Box test to test if the time series is white noise. 
White noise is a time series that is purely random. Below 
there is a test at lag 4.

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Ljung-Box test
Box.test(gdp_ts, lag = 4, type = "Ljung")

```

Ljung-Box test p-value is very small < 0.01, so there is strong evidence that the
time series is not white noise and has seasonality and/or trend.


# Modelling

ARIMA (Auto-regressive integrated moving average) models provide one of the main
approaches to time series forecasting. It is the most widely-used approach to 
time series forecasting, and aim to describe the autocorrelations in the data.

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

# Create the training set
train1 <- window(gdp_ts, end = c(2016, 4))

# Fit a seasonal ARIMA model with lambda = 0 - box cox transformation 
fit <- auto.arima(train1, lambda = 0)

```



The final fitted model was produced by the auto.arima() function of the **forecast library**.
It rapidly estimates a series of model and return the best, according to either AIC, AICc or BIC value.
After fitting the ARIMA model, it is essential to check that the residuals are 
well-behaved (i.e., no outlines or patterns) and resemble white noise. Below there
are some residual plots for the fitted ARIMA model. 

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Plot residuals 
checkresiduals(fit)

```

We can say that the model is fairly good, since the residuals are closely normally 
distributed, have no real pattern and autocorrelations are not significant.  
The final model is a seasonal **ARIMA(2,1,1)(0,1,1)[4]**. Both seasonal and first differences 
have been used, indicated by the middle slot in each part of the model. Also, one lagged error 
and one seasonally lagged error has been selected, indicated by the last slot in 
each part of the model. Two autoregression terms have been used, indicated by the first slot in the 
model. No seasonal autoregression terms have been used. 

Finally, the accuracy of the forecasting model is examined. Below 
there is a test for the model accuracy, using all four quarters in 2017 & the 1st quarter of 2018 
as a test set.

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Test accuracy
accuracy(forecast(fit, h = 5), gdp_ts)

```

MAPE (Mean absolute percentage error) for the test set is 1.68, so we can conclude that the prediction 
accuracy of the model is around **98.3 %**.  

Below there is a time series plot of the Eurozone countries quarterly GDP forecasts
for 2018-19. 

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Fit a seasonal ARIMA model with lambda = 0 - box cox transformation for all data
fit <- auto.arima(gdp_ts, lambda = 0)

# Plot forecasts
fit %>% 
  forecast(h = 8) %>% 
  autoplot() +
  labs(title = "Forecasting of Eurozone countries quarterly GDP",
       subtitle = "In million euro (€), for years 2018-19",
       y = "GDP (€)")+
  scale_x_continuous(breaks = seq(1995, 2020, 1)) +
  theme_fivethirtyeight() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

- It looks that GDP will keep growing in the next couple of years. In particular 
the forecasts for the future quarters are:
```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
fit %>% 
  forecast(h = 7) 
```

- It seems that by the end of 2019, there is a strong possibility that in one or more of 
the quarters of the year, the GDP will break the barrier of **3 trillion €**

More models developed using other forecasting approaches, such as 
**exponential smoothing**(exponential smoothing methods with trend and seasonality Holt-Winters) &
**exponential triple smoothing**, but the ARIMA model performance was better. 

[Full R code](https://github.com/mantoniou/Blog/blob/master/content/post/2018-08-forecasting-arima-gdp-eurozone.Rmd)

