---
title: Predicting the future popularity of programming languages
author: ''
date: '2019-09-14'
description: "By using Stack Overflow usage data, build multiple forecasting models to predict future popularity of programming languages"
slug: predicting-the-future-popularity-of-programming-languages
categories:
  - R
tags:
  - Forecasting
  - Arima
  - Exponential Smoothing
---


# Insert data and ETL

Stack overflow data were used for this analysis. The dataset was downloaded from the [Stack Exchange Data Explorer](https://data.stackexchange.com/). 
The processed file is also downloadable [here](https://gist.github.com/dgrtwo/a30d99baa9b7bfc9f2440b355ddd1f75). 
It was used by [David Robinson](https://twitter.com/drob) in a [datacamp](https://www.datacamp.com/) project.

Each Stack Overflow question has a tag, which marks a question to describe its
topic or technology. This data has one observation for each pair of a tag and a year, showing the 
number of questions asked in that tag in that year and the total number of 
questions asked in that year. For instance, there were 54 questions asked about
the .htaccess tag in 2008, out of a total of 58390 questions in that year.



```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
library(tidyverse)
library(plotly)
library(DT)
library(kableExtra)
library(knitr)

theme_set(theme_minimal())

yearly_tag <- read_csv("https://gist.githubusercontent.com/dgrtwo/a30d99baa9b7bfc9f2440b355ddd1f75/raw/700ab5bb0b5f8f5a14377f5103dbe921d4238216/by_tag_year.csv")

kable(head(yearly_tag)) %>% 
  kable_styling()
```


Instead of using the counts of each tag questions in a year we'll calculate & use the 
**fraction of each tag questions** and the overall questions in that year. It is
more convenient to perform the comparison now.

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Add fraction column
yearly_tag <- 
    yearly_tag %>% 
    mutate(fraction = round(number/year_total, 4))

 
# Print the new table
kable(head(yearly_tag)) %>% 
  kable_styling()
```

# How have popular programming languages changed over time?

It would be interesting to look at the popularity of the top programming
languages. In particular, the following programming languages are included:

- java 
- JavaScript  
- c#  
- c++  
- python  
- php  
- ruby  
- r  

The fraction of each tag questions (on the overall questions in the year) used 
for this comparison.


```{r echo=FALSE, fig.height=7, fig.width=10, message=FALSE, warning=FALSE, paged.print=FALSE}

# Get the six largest tags
programming_lang <- c("r", "python", "c#", "java", "JavaScript", "php", "c++", "ruby")

yearly_top <- 
  yearly_tag %>% 
  filter(tag %in% programming_lang)

d_ends <- 
  yearly_top %>% 
  group_by(tag) %>% 
  slice(n()) %>% 
  pull(fraction)

d_ends[1] <- 0.053
d_ends[6] <- 0.024

d_labels <- 
  yearly_top %>% 
  group_by(tag) %>% 
  slice(n()) %>% 
  pull(tag)



# Filter for the six largest tags
ggplot(yearly_top) +
  geom_line(aes(x = year, y = fraction, color = tag), size = 1.5, alpha = .8) +
  geom_point(aes(x = year, y = fraction, color = tag), size = 2) +
  scale_x_continuous(expand = c(0, 0), breaks = c(2008:2018)) +
  scale_y_continuous(labels = scales::percent, breaks = c(0, .025, .05, .075, .1, .125), sec.axis = sec_axis(~ ., breaks = d_ends, labels = d_labels)) +
  labs(title = "Fraction of total questions per year in Stack Overflow",
       subtitle = "for top programming languages",
       x = "", 
       y = "Fraction of total queries in the year") +
  theme(legend.position = "none")



```

We can identify that some languages are rising & others that falling in popularity. 
In particular JavaScript, java, c#, c++, ruby are falling and python with R 
(languages used in analytics) are rising. Although this is true
we should check if the total number of questions is growing significantly. 


```{r echo=FALSE, fig.height=7, fig.width=10}

yearly_tag %>% 
  group_by(year) %>% 
  summarise(year_total = first(year_total)) %>%
  filter(year <= 2017) %>% 
  ggplot() +
  geom_line(aes(year, year_total), color = "steelblue", size = 1.5, alpha = .5 ) +
  geom_point(aes(year, year_total), color = "steelblue", size = 1.5) +
  scale_x_continuous(breaks = c(2008:2017)) +
  labs(title = "Total number of questions in Stack overflow per year",
       x = "",
       y = "Num. of questions")


```

It is clear that since 2013 the total number of questions is not significantly 
growing. So we can say that the trends of the programming languages could be significant, especially from 2013 onward. 

# Predicting the future popularity of programming languages

It would be interesting to predict the future popularity of the programming
languages. I'll use the [forecast package](https://cran.r-project.org/web/packages/forecast/forecast.pdf) 
to generate predictions. In particular, I'm combining the power of the main forecasting
methodologies, **ARIMA & Exponential smoothing**. In particular for each time-series
(each programming language) 2 separate models are created, using ARIMA* & Exponential smoothing methods**, 
and the best one is selected for prediction. **MAPE (mean absolute percentage error)** 
is chosen to evaluate the forecasting models.   
* ARIMA (Auto-regressive integrating moving average) is a very popular technique for time
series modelling. It describes the correlation between data points and takes into account
the difference of the values.  
** Exponential Smoothing methods include simple exponential smoothing (larger weights are
assigned to more recent observations than to observations from the distant past), 
double exponential smoothing or Holt linear trend model (also takes account the 
trend of the series) and triple exponential smoothing or Host's Winters method 
(also takes account both the trend and the seasonality of the time series)


```{r message=FALSE, include=FALSE, paged.print=FALSE}
library(forecast)
library(sweep)


# Get tags for top programming languages
programming_lang <- c("r", "python", "c#", "java", "JavaScript", "php", "c++", "ruby")

# Create the dataset
yearly_nest <- 
yearly_tag  %>% 
  filter(tag %in% programming_lang) %>%
  arrange(tag, year) %>%
  select(tag, fraction) %>% 
  group_by(tag) %>% 
  nest(.key = "data.tbl") %>% # nest it
  mutate(data.ts = map(.x    = data.tbl, #create ts object
                       .f    = ts,
                       start = 2008)
                       ) %>% 
  mutate(fit_ets = map(data.ts, ets)) %>%
  mutate(summary_ets = map(fit_ets, summary)) %>%
  mutate(mape_ets = map(summary_ets, 5)) %>% 
  mutate(fit.arima = map(data.ts, auto.arima)) %>%
  mutate(summary_arima = map(fit.arima, summary)) %>%
  mutate(mape_arima = map(summary_arima, 5)) %>% 
  
  mutate(final_model = if_else(as.numeric(mape_arima) <= as.numeric(mape_ets), fit.arima, fit_ets)) %>% 
  
  mutate(predict = map(final_model, forecast, h = 5)) %>% 
  mutate(sweep = map(predict, sw_sweep)) %>% 
  unnest(sweep) %>% 
  mutate(fraction = if_else(fraction < 0, 0, fraction))


table_a <- 
yearly_tag  %>% 
  filter(tag %in% programming_lang) %>%
  arrange(tag, year) %>%
  select(tag, fraction) %>% 
  group_by(tag) %>% 
  nest(.key = "data.tbl") %>% # nest it
  mutate(data.ts = map(.x    = data.tbl, #create ts object
                       .f    = ts,
                       start = 2008)
                       ) %>% 
  mutate(fit_ets = map(data.ts, ets)) %>%
  mutate(summary_ets = map(fit_ets, summary)) %>%
  mutate(mape_ets = map(summary_ets, 5)) %>% 
  mutate(fit.arima = map(data.ts, auto.arima)) %>%
  mutate(summary_arima = map(fit.arima, summary)) %>%
  mutate(mape_arima = map(summary_arima, 5)) %>% 
  select(tag, mape_arima, mape_ets) %>% 
  mutate(mape_arima = round(as.numeric(mape_arima), 2), 
         mape_ets = round(as.numeric(mape_ets), 2)) 


```


```{r echo=FALSE}
table_a %>% 
  kable() %>% 
  kable_styling()

```

As you can see above, R & C++ predictions are better when applying exponential 
smoothing method than ARIMA. For the rest of the programming languages, ARIMA seems
to be the best methodology. 

Below there is a table with the future predictions, using the best performing 
model


```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

filter(yearly_nest, key == "forecast") %>%
  top_n(15) %>% 
  kable() %>% 
  kable_styling()

```

Below there is a plot with the future predictions


```{r echo=FALSE, fig.height=7, fig.width=10, message=FALSE, warning=FALSE, paged.print=FALSE}

d_ends <- yearly_nest %>% 
  group_by(tag) %>% 
  slice(n()) %>% 
  pull(fraction)

d_ends[4] <- 0.005

d_labels <- yearly_nest %>% 
  group_by(tag) %>% 
  slice(n()) %>% 
  pull(tag)


# Create the plot 
yearly_nest %>% 
  ggplot() +
  theme_minimal() +
  geom_line(aes(x = index, y = fraction, color = tag), size = 1.5, alpha = .8) +
  geom_point(aes(x = index, y = fraction, color = tag, shape = key ), size = 2) +
  scale_x_continuous(expand = c(0, 0), breaks = c(2008:2024)) +
  geom_rect(data=data.frame(xmin = 2018, xmax = Inf, ymin = -Inf, ymax = Inf),
    aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax), fill="steelblue", alpha=0.2) +
  geom_text(aes(x = 2019, y = 0.15, label = "Prediction", fill = 1), nudge_x = 1.5, colour = "white", size = 5) +
  scale_y_continuous(labels = scales::percent,  sec.axis = sec_axis(~ ., breaks = d_ends, labels = d_labels)) +
  labs(title = "Predicting future fraction of total questions per year in Stack Overflow",
       subtitle = "for top programming languages",
       x = "", 
       y = "Fraction of total queries in the year") +
  theme(legend.position = "none") 


```


# Summary

To compare programming languages popularity we used the fraction of total questions
(on Stack Overflow) that concern each language for the last 10 years. 
Stack overflow is, by far, the most popular platform for 
questions and answers on a wide range of topics in computer programming.

Of course, there are a lot of other metrics that could be used to measure
popularity. Furthermore, forecasting models, like all models, are not perfect and could have higher deviations than predicted. 
But, given the relatively small MAPE for all time series (less than 9 % for all), 
the predictions should be a good indication for the future popularity of programming languages.  

Overall, the results are the following:  
- Analytics programming languages **(Python & R)** will continue gaining popularity  
- **Java** will gain a little and then keep a constant popularity  
- **JavaScript, C# & C++** will loose significant popularity  
- **PHP & Ruby** could lose almost all their popularity and become obsolete in the next 5 years


