---
title: Clustering of EU countries
author: ''
date: '2018-10-27'
description: "Segmentation of EU countries, including cluster analysis based on consumer & business confidence indicators"

slug: clustering-of-eu-countries
categories:
  - R
tags:
  - clustering
  - K-means
  - PCA
---



In this article I used a public Eurostat dataset, to develop a 
segmentation of the EU countries. 
This dataset consists of 5 confidence indicators:  

- **Consumer Confidence Indicator**  
- **Construction confidence indicator**  
- **Industrial confidence indicator**  
- **Retail confidence indicator**  
- **Services Confidence Indicator**  

These indicators are formed via qualitative surveys which are conducted on a monthly 
basis in the following areas:
**manufacturing industry, construction, consumers, retail trade, services and financial services**. 
These surveys started in 1980 and gradually include all the new EU 
members. About 137,000 firms and more than 41,000 consumers are currently surveyed
every month across the EU. 

The used metrics is the balance i.e. the difference between positive and negative
answers (in percentage points of total answers), as index, as confidence indicators
(arithmetic average of balances).


More information about these surveys can be found at this 
[link](https://ec.europa.eu/eurostat/cache/metadata/en/ei_bcs_esms.htm)

The eurozone consists of 19 countries: Austria, Belgium, Cyprus, 
Estonia, Finland, France, Germany, Greece, Ireland, Italy, Latvia, Lithuania, 
Luxembourg, Malta, the Netherlands, Portugal, Slovakia, Slovenia, and Spain. 

The [Eurostat package](https://cran.r-project.org/web/packages/eurostat/index.html) 
used to obtain the original datasets.

More details about the ETL steps can be found, in the actual code, at the link 
at the end of the article.

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}

# Libraries
library(tidyverse)
library(rvest)
library(eurostat)


```



```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
# Eurostat Confidence Indicators dataset #######################################
dat <- get_eurostat("ei_bssi_m_r2", time_format = "date")
dat <- label_eurostat(dat)

sent <- 
  dat %>% 
  filter(s_adj == "Unadjusted data (i.e. neither seasonally adjusted nor calendar adjusted data)") %>% 
  select(geo, time, indic, values) %>% 
  spread(indic, values) %>% 
  rename(date = time) 

# Save the data
saveRDS(sent, file = "data/tidy/sent.RDS")
```



```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}

# Load the stored dataset
sent <- readRDS("/Users/manos/OneDrive/Projects/R/All_Projects/2018_10_ClusterAnalysis/data/tidy/sent.RDS") 
# Create the final dataset
final <-
  sent %>%
  filter(date >= '2014-01-01') %>%
  select(-date) %>%
  group_by(geo) %>%
  summarise_all(median, na.rm = TRUE) %>% 
  mutate(Business = (`Construction confidence indicator` + `Industrial confidence indicator` +
           `Retail confidence indicator` + `Services Confidence Indicator`)/4,
         Consumers = `Consumer confidence indicator`) %>% 
  mutate(geo = recode(geo, 
                      "Germany (until 1990 former territory of the FRG)" = "Germany",
                      "Czech Republic" = "Czech Rep."))


```


# ETL & Exploratory Analysis

The original dataset consists of 11,340 observations that include these indicators
for EU countries on a monthly basis. In the processed dataset I used i) observations
from 2014 onward & ii) the median values of each variable for each state, so finally the dataset consists of 28 observations in total. One more 
variable was created, by averaging all Business related confidence indicators 
(Construction, Industrial, Retail & Services Confidence Indicators), to use a general **business confidence indicator**.



```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

# Create Map data ##############################################################

library(grid)
library(rworldmap)


# Get the world map
worldMap <- getMap()

# Member States of the European Union
europeanUnion <- c("Austria","Belgium","Bulgaria","Croatia","Cyprus",
                   "Czech Rep.","Denmark","Estonia","Finland","France",
                   "Germany","Greece","Hungary","Ireland","Italy","Latvia",
                   "Lithuania","Luxembourg","Malta","Netherlands","Poland",
                   "Portugal","Romania","Slovakia","Slovenia","Spain",
                   "Sweden","United Kingdom")

# Select only the index of states member of the E.U.
indEU <- which(worldMap$NAME%in%europeanUnion)


# Extract longitude and latitude border's coordinates of members states of E.U. 
europeCoords <- lapply(indEU, function(i){
  df <- data.frame(worldMap@polygons[[i]]@Polygons[[1]]@coords)
  df$region =as.character(worldMap$NAME[i])
  colnames(df) <- list("long", "lat", "region")
  return(df)
})

europeCoords <- do.call("rbind", europeCoords)

```


It would be interesting to see the map plots of the confidence indicators. Below
there are plots with the consumer & averaging business confidence indicators. 

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Create table
europeanUnionTable <- data.frame(country = final$geo, value = final$Business)
europeCoords$value <- europeanUnionTable$value[match(europeCoords$region,europeanUnionTable$country)]


# Produce spatial visualisation
ggplot() + 
  geom_polygon(data = europeCoords, aes(x = long, y = lat, group = region, fill = value),
                             colour = "black", size = 0.1) +
  coord_map(xlim = c(-13, 35),  ylim = c(32, 71)) + 
  scale_fill_gradient(name = "Business Confidence Indicator", low = "#FF0000FF", high = "#FFFF00FF", na.value = "grey50") +
  theme(axis.text.x = element_blank(),
  axis.text.y = element_blank(), axis.ticks.x = element_blank(),
  axis.ticks.y = element_blank(), axis.title = element_blank(),
  plot.margin = unit(0 * c(-1.5, -1.5, -1.5, -1.5), "lines")) + 
  scale_fill_gradient(name = "Business \nConfidence", low = "#FF0000FF", high = "#FFFF00FF", na.value = "grey50") +
  labs(title = "Business Confidence per EU country",
       subtitle = "-Median value of aggregated bus. conf. indicators for 2014-2018 \n-Red indicates low confidence & yellow indicates high confidence levels")+
  theme(panel.grid = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_blank())

```


```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

# Create table
europeanUnionTable <- data.frame(country = final$geo, value = final$Consumers)
europeCoords$value <- europeanUnionTable$value[match(europeCoords$region,europeanUnionTable$country)]


# Produce spatial visualisation
ggplot() + 
  geom_polygon(data = europeCoords, aes(x = long, y = lat, group = region, fill = value),
                             colour = "black", size = 0.1) +
  coord_map(xlim = c(-13, 35),  ylim = c(32, 71)) + 
  scale_fill_gradient(name = "Consumers Confidence Indicator", low = "#FF0000FF", high = "#FFFF00FF", na.value = "grey50") +
  theme(axis.text.x = element_blank(),
  axis.text.y = element_blank(), axis.ticks.x = element_blank(),
  axis.ticks.y = element_blank(), axis.title = element_blank(),
  plot.margin = unit(0 * c(-1.5, -1.5, -1.5, -1.5), "lines")) + 
  scale_fill_gradient(name = "Consumer \nConfidence", low = "#FF0000FF", high = "#FFFF00FF", na.value = "grey50") +
  labs(title = "Consumers Confidence per EU country",
       subtitle = "-Median value of consumer conf. indicator for 2014-2018 \n-Red indicates low confidence & yellow indicates high confidence levels")+
  theme(panel.grid = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_blank())


```

- It is clear that there are significant differences between countries.  
- Northern Europe countries tend to have higher consumers confidence 
indicators.  
- There are some outliers here. E.g. Greece has significantly lower confidence indicator
than the rest of the nations. 


Below there is a scatterplot with marker labels indicating the positioning of
each country in respect to consumer and business confidence indicators. 

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(ggthemes)

final <- filter(final, geo %in% europeanUnion) 

# Scatter plot
final %>% 
ggplot(aes(Consumers, Business)) +
  geom_point(col = "red")+
  geom_text(aes(label = geo), check_overlap = TRUE, vjust = 1, hjust = 1) +
  theme_fivethirtyeight() +
  theme(axis.title = element_text()) +
  labs(x = "Consumer Confidence",
       y = "Business Confidence",
       title = "Scatterplot of EU countries confidence levels",
       subtitle = "Data from 2014-2018") +
  xlim(-65, 20)

```

We can indicate some patterns from the plot above, similar to the findings before:   
- There are some outliers, such as Greece(bottom left) & Sweden(top right)  
- There is a group of countries that are placed in the middle of the plot, indicating
average consumer & business confidence  
- In all countries, the business confidence is substantially higher than
consumer confidence


# Segmentation

Since there are quite a few differences between countries regarding 
confidence indicators, it would be interesting to develop a segmentation, to check how well the countries are forming teams.  
The **k-means** algorithm used for the segmentation. 
It is the widest used unsupervised learning algorithm. The procedure follows
a simple and easy way to classify a given data set through a certain number of 
clusters (assume k clusters) fixed apriori. The main idea is to define k centers,
one for each cluster. These centers should be placed in a cunning way because of
different location causes different result. So, the better choice is to place
them as much as possible far away from each other. The next step is to take each 
point belonging to a given data set and associate it to the nearest center. When 
no point is pending, the first step is completed and an early group age is done. 
At this point we need to re-calculate k new centroids as barycenter of the clusters
resulting from the previous step. After we have these k new centroids, a new 
binding has to be done between the same data set points and the nearest new center.
A loop has been generated. As a result of this loop we may notice that the k 
centers change their location step by step until no more changes are done or in 
other words centers do not move any more.


## Indicate suitable number of clusters

The elbow (scree) plot below, is used to check for the suitable number of clusters. 
So what we are looking for, is the point at which the curve in the plot begins
to flatten out. 


In detail, the total within cluster sum of squares is calculated (the sum of 
euclidean distances between each observation and the centroid corresponding to 
the cluster to which the observation is assigned). 


```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(purrr)
library(cluster)

# Produce matrix of distances
kmeans.dat <- dist(scale(final[, 2:6]))

# Use map_dbl to run many models with varying value of k (centers)
tot_withinss <- map_dbl(1:10,  function(k){
  model <- kmeans(x = kmeans.dat, centers = k)
  model$tot.withinss
})

# Generate a data frame containing both k and tot_withinss
elbow_df <- data.frame(
  k = 1:10,
  tot_withinss = tot_withinss
)

# Elbow plot
ggplot(elbow_df, aes(x = k, y = tot_withinss)) +
  geom_line() +
  scale_x_continuous(breaks = 1:10) +
  theme_fivethirtyeight() +
  theme(axis.title = element_text()) +
  labs(y = "Total within sum of squares",
       x = "Number of clusters",
       title = "Elbow (scree) plot",
       subtitle = "Check optimal number of clusters")

```


**Silhouette analysis**

In general, silhouette analysis determines how well each of the observations fit into 
corresponding cluster (Higher values are better).

It involves calculating a measurement called the silhouette width for every 
observation:  
- A value close to 1 suggests that this observation is well matched to its current cluster.  
- A value of 0 suggests that it is on the border between two clusters and possibly 
belong to either one.  
- A value close to -1 indicates that the observation has a better fit to its 
closest neighbouring cluster.



```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

library(purrr)
library(cluster)

# Use map_dbl to run many models with varying value of k
sil_width <- map_dbl(2:10,  function(k){
  model <- pam(x = kmeans.dat, k = k)
  model$silinfo$avg.width
})

# Generate a data frame containing both k and sil_width
sil_df <- data.frame(
  k = 2:10,
  sil_width = sil_width
)

# Plot the relationship between k and sil_width
ggplot(sil_df, aes(x = k, y = sil_width)) +
  geom_line() +
  scale_x_continuous(breaks = 2:10)+
  theme_fivethirtyeight() +
  theme(axis.title = element_text()) +
  labs(y = "Average Silhouette width",
       x = "Number of clusters",
       title = "Silhouette Analysis")

```

In conclusion, I proceed with 2 clusters, as it is more suitable both from a 
technical & practical point of view.


Below there is a table plot with **information about all clusters**, indicating the 
mean value of each confidence indicator within the cluster


```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
set.seed(42)

# Build a k-means model with a k of 2
model_customers <- kmeans(kmeans.dat, centers = 2)

# Extract the vector of cluster assignments from the model
clust_countries <- model_customers$cluster

# Build the dataframe by adding the cluster
segment_countries <-
  mutate(final[1:6], cluster = clust_countries) %>%
  mutate(cluster = recode(cluster,
                             "1" = "B",
                             "2" = "A"))

library(kableExtra)

# Calculate the mean for each variable & print the table
segment_countries %>%
  group_by(cluster) %>%
  add_tally() %>%
  summarise_each(funs(round(mean(., na.rm = TRUE),2))) %>%
  select("cluster", "n", 3:ncol(final)) %>% 
  rename("Construction"  =  "Construction confidence indicator",
         "Consumer"  =  "Consumer confidence indicator",
         "Industrial"  =  "Industrial confidence indicator",
         "Retail"  =  "Retail confidence indicator",
         "Services"  =  "Services Confidence Indicator"
         ) %>% 
  kable("html", caption = 'Confidence indicators') %>%
  kable_styling(position = "center") %>% 
  scroll_box(width = "100%", height = "200px")

  # kable() %>%
  # kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

```


- Cluster A tend to have higher values than cluster B on Construction confidence indicator, Consumer confidence indicator & Services Confidence Indicator

- Both clusters tend to have similar values on Industrial confidence indicator & Retail confidence indicator

## Principal components
It would be nice to plot these clustering results and check these out visually. 
But it is impossible to visualise so many variables, as various dimensions are
required. 
One way to overcome this is to use some dimensionality reduction technique. In particular, PCA (principal components analysis) is used.  
It finds structure in features and aid in visualization.  
In particular:  
- It will find some linear combination of variables to create principal components (new features)    
- Maintain most variance in the data    
- Principal components (new features) are uncorrelated (i.e. orthogonal to each other)

The plot below (bi-plot) shows all the original observations plotted on the first
two principal components. 
        
```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

# apply pca
pca_soft <- prcomp(scale(kmeans.dat))

library(ggord)

# Print biplot
ggord(pca_soft, segment_countries$cluster, ellipse = FALSE, hull = TRUE, arrow = 1, vec_ext = 3, veccol = 'red', veclsz = 1) +
  labs(title = "Biplot of principal components analysis",
       subtitle = "Check how well the clusters are separated")

```

- The A cluster clearly stands out from cluster B
- There is just a minor overlap 


```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

# Create table with results
europeanUnionTable <- data.frame(country = segment_countries$geo, value = factor(segment_countries$cluster))
europeCoords$value <- europeanUnionTable$value[match(europeCoords$region,europeanUnionTable$country)]


# Plot the map with clusters
ggplot() + 
  geom_polygon(data = europeCoords, aes(x = long, y = lat, group = region, fill = value),
                             colour = "black", size = 0.1) +
  coord_map(xlim = c(-13, 35),  ylim = c(32, 71))+ 
  theme(
  axis.text.x = element_blank(),
  axis.text.y = element_blank(), axis.ticks.x = element_blank(),
  axis.ticks.y = element_blank(), axis.title = element_blank(),
  plot.margin = unit(0 * c(-1.5, -1.5, -1.5, -1.5), "lines")) + 
  labs(title = "EU clusters based on confidence indicators",
       subtitle = "- Green indicates higher & red lower conf. indicators",
       fill = "Cluster")+
  theme(panel.grid = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_blank()) +
   scale_fill_manual(values=c("#33A02C", "#E31A1C"))

```

- In general, northern Europe countries tend to have higher confidence indicators that place them in cluster A  
- The clustering results can confirm the theory of the two-speed EU (significant differences 
between countries in many aspects as the standard of living etc.) 


Below there is a table of all EU countries indicating their cluster and the median
confidence indicators (since 2014)

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

library(DT)

# Print the table
segment_countries %>% 
  select(c(1, "cluster", 2:6)) %>% 
  mutate(cluster = factor(cluster)) %>% 
  rename(Cluster = cluster, Country = geo) %>% 
  rename("Construction"  =  "Construction confidence indicator",
         "Consumer"  =  "Consumer confidence indicator",
         "Industrial"  =  "Industrial confidence indicator",
         "Retail"  =  "Retail confidence indicator",
         "Services"  =  "Services Confidence Indicator"
         ) %>% 
  arrange(Cluster, Country) %>% 
  kable("html", caption = '') %>%
  kable_styling(position = "center") %>% 
  scroll_box(width = "100%", height = "300px")
# 
#   datatable(filter = 'top', options = list(pageLength = 10, autoWidth = TRUE, dom = 'pt', scrollX = TRUE))

```

[Full R code](https://github.com/mantoniou/Blog/blob/master/content/post/2018-10-27-clustering-of-eu-countries.Rmd)

